Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-12-29 manuscript/manuscript.tex
Remark that D=2 would be ok for gz at 260km
Edits to exponential density case
Fixes some inconsistencies between the text and figure caption. Move
some descriptions of the different functions for varying b further down
to where they are needed.

\begin{summary}
We present a new methodology to compute the gravitational fields generated by
tesseroids (spherical prisms) whose density varies with depth according to
an arbitrary continuous function.
It approximates the gravitational fields through the Gauss-Legendre Quadrature along
with two discretization algorithms that automatically control its accuracy by adaptively
dividing the tesseroid into smaller ones.
The first one is a preexisting two dimensional adaptive discretization algorithm that
reduces the errors due to the distance between the tesseroid and the computation point.
The second is a new density-based discretization algorithm that
decreases the errors introduced by the variation of the density function with depth.
The amount of divisions made by each algorithm is indirectly controlled
by two parameters: the distance-size ratio and the delta ratio.
We have obtained analytical solutions for a spherical shell with radially variable
density and compared them to the results of the numerical model for linear,
exponential and sinusoidal density functions.
These comparisons allowed us to obtain optimum values for the distance-size and
delta ratios that yield an accuracy of 0.1\% of the analytical solutions.
The resulting optimal values of distance-size ratio for the gravitational potential and
its gradient are 1 and 2.5, respectively.
The density-based discretization algorithm produces no discretizations in the linear
density case, but a delta ratio of 0.1 and 0.01 is needed for the exponential and the
sinusoidal density function, respectively.
Lastly, we apply this new methodology to model the Neuqu\'en Basin, a foreland basin in
Argentina with a maximum depth of over 5000~m, using an exponential density function.
\end{summary}

@@

We observe from Fig.~\ref{fig:D-linear} that the relative errors for the
potential and $g_z$ fall below the 0.1\% threshold at
$D=1$ and $D=2.5$, respectively.
These values are consistent for all shell models and computation grids.

@@

For an exponential density function, the density-based discretization will be applied
before the adaptive discretization algorithm.
This means that optimal values for both the distance-size ratio $D$ and the delta ratio
$\delta$ must be determined.
We perform an error analysis similar to what was done for the linear density case.
But, now the spherical shell will have an exponential density function that assumes the
values of $\rho_\text{out} = 2670$~kg/m$^3$ and $\rho_\text{in} = 3300$~kg/m$^3$ on the
outer and inner surfaces, respectively, defined as follows:

\begin{equation}
    \rho(r') = A e^{- \frac{b}{T} (r' - R_1)} + C,
    \label{eq:density-exp}
\end{equation}

@@

\noindent $T$ is the thickness of the shell ($T = R_2 - R_1$) and $b$ is a dimensionless
constant that determines the variability of the function: a higher value of $b$
increases the maximum slope of the density function.
On exponential density analysis we will use values of $b$ equal to 1, 2, 5, 10, 30
and 100, in order to cover different kind of exponential density variations, from smooth
to sharp (Fig.~\ref{fig:exp-densities}).

@@

We aim to find a combination of the $D$ and $\delta$ that produces a numerical error
lower than the 0.1\% threshold while minimizing computation time.
We use a grid search method and compute the numerical error for every ($D$, $\delta$)
pair belonging to a grid on the $D$-$\delta$ space (Fig.~\ref{fig:grid-search}).
For optimum algorithm performance, we search for the ($D$, $\delta$) pair that minimize
the number of splittings while keeping a numerical error under the 0.1\% threshold,
i.e.~the smallest possible value of $D$ and the highest possible value of $\delta$.

We compute the relative difference between the numerical and analytical results for the
gravitational potential ($V$) and the vertical component of its gradient ($g_z$) on the
computation grids detailed in Table~\ref{tab:grids}.
Because this is a time consuming computation, we limit the analysis only to a single
value of $b$ equal to 30 (Fig.~\ref{fig:exp-densities}).
For the sake of brevity, we compute the maximum differences between the results for each
shell model.
The results are shown in Fig.~\ref{fig:grid-search}, in which points inside the dotted
lines are the ones that present a numerical error lower than the 0.1\% threshold.
Also shown in Fig.~\ref{fig:grid-search} are the $D$ values determined for the linear
density function in the previous section ($D_\text{linear}$).

The minimum values of $D$ that are within the 0.1\% threshold are the $D_\text{linear}$
ones, i.e.~they are the optimal values of the distance-size ratios for $V$ and $g_z$.
These results indicate that the values of $D_\text{linear}$ can be safely extrapolated
to non linear density cases.
This conclusion can be understood considering that the two dimensional adaptive
discretization and the density-based algorithms are independent of each other: the first
one splits the tesseroid only in the latitudinal and longitudinal directions, while the
latter only does it in the radial direction.
This leads to the fact that the value of $\delta$ needed to limit the numerical error
due to the density variation is independent of the value of $D$ and vice versa.
That's why the optimal distance-size ratios for the exponential density case is the same
as the ones for the linear density case.

Because the grid search was limited to specific values of $b$ and used only one grid, we
will perform a more detailed analysis for $\delta$ determination in case of exponential
density tesseroids.

@@

Having chosen values of $D$ equal to the ones obtained for the linear density case, we
are free to explore the integration error as a function of $\delta$ in more detail and
how it varies for different values of $b$ (Eq.~\ref{eq:density-exp}).
Now we perform the error calculations for values of delta ranging from $10^{-3}$ to
$10^{1}$, for every grid detailed in Table~\ref{tab:grids} and for the spherical shell
models detailed in Table~\ref{tab:shell-models}.
The calculations are repeated for distinct values of $b$ (1, 2, 5, 10, 30, 100) to
examine the accuracy of the method for different sharpness of the density function.
Because larger $\delta$ values result in fewer tesseroid divisions,
our intention is to find the highest value of $\delta$ whose numerical error is bellow
the $0.1\%$ threshold.

Fig.~\ref{fig:exp-densities} shows the density functions for each value of $b$ and
Fig.~\ref{fig:delta-exponential} shows the resulting relative
differences ($V$ and $g_z$).
In order to improve the visualization of the results, each curve corresponds to the
maximum difference between all shell models for the same value of $b$.
The relative difference for $V$ and $g_z$ falls below the 0.1\% threshold
for each computation grid, shell model and $b$ value if $\delta$ is lower than 0.1.
This results imply that the optimal value of $\delta$ is 0.1 in case of exponential
density.
They also show that there is no significant relation between the $b$ value and the
numerical error, proving that the density-based discretization algorithm is capable of
limiting the error due to the density variation regardless the variation of the
exponential density.

@@

So far we have tested the density-based discretization algorithm against linear and
exponential density functions.
Nevertheless, this new algorithm is well suited to accurately reproduce the
gravitational fields of tesseroids with more complex continuous functions, like
non-monotonic or with one or more inflection points.
Although such density functions are very rare (or maybe even non-existent) between
geological structures, we want to subject the density-based algorithm to such cases in
order to show that it can solve more complex situations other than linear and
exponential functions.

@@

Because there is no direct relation between the values assigned to $D$ and $\delta$ and
the error of the computation, we empirically determined the optimum values for these
parameters.
These values minimize the computational load while maintaining the numerical error below
0.1\% of an analytical solution.
The density functions used to establish the optimum values were a linear, an exponential
and a sinusoidal function.
The linear function represents the smoothest variation of density and does not require
density-based discretization at all.
We analysed the error for exponential functions ranging from smooth to sharp variations
in depth to test the accuracy of the density-based discretization.
And the error for sinusoidal functions with different wavelengths.
Through this analysis, we have obtained optimal values for the distance-size ratio of
1 and 2.5 for the potential an its gradient, respectively.
While values of $\delta$ equal to 0.1 and 0.01 are needed in order to guarantee a 0.1\%
accuracy for an exponential and a sinusoidal density, respectively.
These results could be extrapolated: a $\delta$ value equal to 0.01 is needed in order
to achieve the accuracy threshold in case of an arbitrary continuous density function.
Although we encourage to perform similar accuracy tests for functions other than the
ones we studied, this results could be safely generalized to other common depth-density
relations among geophysical applications, like quadratic functions, without loss of
accuracy.

>>>>

\begin{summary}
We present a new methodology to compute the gravitational fields generated by
tesseroids (spherical prisms) whose density varies with depth according to
an arbitrary continuous function.
It approximates the gravitational fields through the Gauss-Legendre Quadrature along
with two discretization algorithms that automatically control its accuracy by adaptively
dividing the tesseroid into smaller ones.
The first one is a preexisting two dimensional adaptive discretization algorithm that
reduces the errors due to the distance between the tesseroid and the computation point.
The second is a new density-based discretization algorithm that
decreases the errors introduced by the variation of the density function with depth.
The amount of divisions made by each algorithm is indirectly controlled
by two parameters: the distance-size ratio and the delta ratio.
We have obtained analytical solutions for a spherical shell with radially variable
density and compared them to the results of the numerical model for linear,
exponential and sinusoidal density functions.
These comparisons allowed us to obtain optimal values for the distance-size and
delta ratios that yield an accuracy of 0.1\% of the analytical solutions.
The resulting optimal values of distance-size ratio for the gravitational potential and
its gradient are 1 and 2.5, respectively.
The density-based discretization algorithm produces no discretizations in the linear
density case, but a delta ratio of 0.1 and 0.01 is needed for the exponential and the
sinusoidal density function, respectively.
Lastly, we apply this new methodology to model the Neuqu\'en Basin, a foreland basin in
Argentina with a maximum depth of over 5000~m, using an exponential density function.
\end{summary}

@@

We observe from Fig.~\ref{fig:D-linear} that the relative errors for the potential and
$g_z$ fall below the 0.1\% threshold at $D=1$ and $D=2.5$, respectively.
Notably, a value of $D=2$ would suffice for $g_z$ in the case of the satellite height
grid.
For all other configurations, these values are consistent and independent of the shell
model thickness or geographic location.

@@

For an exponential density function, the density-based discretization will be applied
before the adaptive discretization algorithm.
This means that optimal values for both the distance-size ratio $D$ and the delta ratio
$\delta$ (Eq.~\ref{eq:delta-density}) must be determined.
We perform an error analysis similar to what was done for the linear density case.
Now the spherical shell will have an exponential density function that assumes the
values of $\rho_\text{out} = 2670$~kg/m$^3$ and $\rho_\text{in} = 3300$~kg/m$^3$ on the
outer and inner surfaces, respectively, defined as follows:

\begin{equation}
    \rho(r') = A e^{- \frac{b}{R_2 - R_1} (r' - R_1)} + C,
    \label{eq:density-exp}
\end{equation}

@@

\noindent and $b$ is a dimensionless
constant that determines the variability of the function.
A higher value of $b$ increases the maximum slope of the density function
(Fig.~\ref{fig:exp-densities}).

@@

We aim to find a combination of the $D$ and $\delta$ that produces a numerical error
lower than the 0.1\% threshold while minimizing computation time.
We use a grid search method and compute the numerical error for every ($D$, $\delta$)
pair belonging to a grid on the $D$-$\delta$ space (Fig.~\ref{fig:grid-search}).
For optimum algorithm performance, we search for the ($D$, $\delta$) pair that minimizes
the number of tesseroid division while keeping a numerical error under the 0.1\%
threshold.
This requirement translates into the smallest possible value of $D$ and the highest
possible value of $\delta$.

Because this is a time consuming computation, we limit the analysis to a high value of
$b=30$ (Fig.\ref{fig:exp-densities}) and the global grid defined in
Table~\ref{tab:grids}.
We compute the relative absolute difference between the numerical and analytical results
of the gravitational potential ($V$) and its vertical derivative ($g_z$) for all shell
models defined in Table~\ref{tab:shell-models}.
For the sake of brevity, Fig.~\ref{fig:grid-search} shows the maximum difference values
obtained from all shell models.
The dotted lines in Fig.~\ref{fig:grid-search} represent a contour of 0.1\% relative
error.
Points inside the dotted are within the acceptable threshold.
Also shown in Fig.~\ref{fig:grid-search} are the $D$ values determined for the linear
density function in the previous section ($D_\text{linear}$).

The smallest values of $D$ that are within the 0.1\% threshold coincide with
$D_\text{linear}$ for both $V$ and $g_z$.
These results indicate that $D_\text{linear}$ can be safely extrapolated to non-linear
density cases.
This is not surprising considering that the two dimensional adaptive
discretization and the density-based discretization are independent of each other.
The former divides the tesseroid in the horizontal dimensions, while the latter only
divides in the radial dimension.
Hence, the optimal value of $\delta$ is likely to be independent of the optimal value of
$D$.
Because the grid search was limited to a specific computation grid and value of $b$,
we perform a more detailed analysis in the following section to determine an optimal
value of $\delta$ for the exponential density case.

@@

Having chosen values of $D$ equal to the ones obtained for the linear density case, we
are free to explore the integration error as a function of $\delta$ in more detail.
We will compute the difference between the numerical and analytical results for all
combinations of computation grid (Table~\ref{tab:grids}) and spherical shell model
(Table~\ref{tab:shell-models}), varying $\delta$ from $10^{-3}$ to $10^{1}$.
The calculations are repeated for each $b \in \{1, 2, 5, 10, 30, 100\}$
(Fig.~\ref{fig:exp-densities}) to examine the accuracy of the method for density
functions of different sharpness.
Because larger $\delta$ values result in fewer tesseroid divisions,
our intention is to find the highest value of $\delta$ that produces a relative
difference bellow the $0.1\%$ threshold.

Fig.~\ref{fig:delta-exponential} shows the resulting relative
absolute differences for $V$ and $g_z$ as a function of $\delta$.
For the sake of brevity, each curve corresponds to the maximum difference between all
shell models.
The curves for $b=1$ and $b=2$ are below the 0.1\% threshold for all values of $\delta$
and do not change for $\delta > 0.8$, indicating that the density
functions are sufficiently smooth and no density discretizations are necessary.
For all other values of $b$, the difference falls below the 0.1\% threshold
if $\delta = 0.1$.
These results indicate that there is no significant relationship between the
sharpness of the exponential density function and the numerical error.

@@

So far we have tested the density-based discretization algorithm against linear and
exponential density functions.
Nevertheless, this new algorithm is well suited to more complex continuous functions,
for example non-monotonic functions or those with multiple inflection points.
Although such density functions are rarely, if ever, seen in geological structures,
we want to subject our algorithm to such cases in order to show that it can solve more
complex situations than linear and exponential functions.

@@

Because there is no direct relation between the values assigned to $D$ and $\delta$ and
the error of the computation, we empirically determined the optimal values for these
parameters.
These values minimize the computational load while maintaining the numerical error below
0.1\% of an analytical solution.
The density functions used to establish the optimal values were a linear, an exponential
and a sinusoidal function.
The linear function represents the smoothest variation of density and does not require
density-based discretization at all.
We analysed the error for exponential functions ranging from smooth to sharp variations
in depth to test the accuracy of the density-based discretization.
And the error for sinusoidal functions with different wavelengths.
Through this analysis, we have obtained optimal values for the distance-size ratio of
1 and 2.5 for the potential an its gradient, respectively.
While values of $\delta$ equal to 0.1 and 0.01 are needed in order to guarantee a 0.1\%
accuracy for an exponential and a sinusoidal density, respectively.
These results could be extrapolated: a $\delta$ value equal to 0.01 is needed in order
to achieve the accuracy threshold in case of an arbitrary continuous density function.
Although we encourage to perform similar accuracy tests for functions other than the
ones we studied, this results could be safely generalized to other common depth-density
relations among geophysical applications, like quadratic functions, without loss of
accuracy.
